\vspace{2cm} \LARGE \bf Seasonal influenza circulation patterns and projections for 2017-2018

### 2017-09-19 auspice/reports/tex/sep-2017.tex
Revise report

\clearpage
\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree.png}
  \caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.} Colors are chosen to highlight clades 1--5. Clades are in reverse order, that is clade 5 is on top, clade 1 is at the bottom. Each of these clades is non-nested and so all compete against each other.
  }
  \label{h3n2_tree}
\end{figure}

We observe substantial diversification within the 3c2.a clade over the
past year. There are 5 major subclades within 3c2.a that have reached
appreciable frequency (\FIG{h3n2_tree}). These clades are characterized by the following
mutations and highlighed in the phylogenetic tree in \FIG{h3n2_tree}:

@@

All of these clades have risen in frequency and are now competing with
one another (\FIG{h3n2_mutations}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
terms of sequence, but is at low frequency globally (currently less than
1\%) despite making up the a substantial fraction of 2017 Oceania
viruses. Clade 2 (121K, 144K) had a large upswing around Dec 2016, but has since declined slightly to $\sim$21\% global frequency. Clade 3 (131K, 142K,
261Q) and clade 4 (135K, HA2:150E) have each slowly climbed to
approximately 20\% frequency. However, clade 5 (92R, 311Q) has recently
increased to 39\% in global frequency.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/h3n2_mutations.png}
  \caption{\textbf{Frequency trajectories of 3c2.a subclades.}
  We estimate frequencies of different clades based on sample counts and collection dates.
  Each clade is estimated according to a characteristic mutation unique to this clade.
  These estimates are based on all available data and global frequencies are weighted according to regional population size and relative sampling intensity .
  We use a Brownian motion process prior to smooth frequencies from month-to-month.
  Transparent bands show an estimate the 95\% confidence interval based on sample counts.
  }
  \label{h3n2_mutations}
\end{figure}

@@

Measuring clade frequency growth and decline over the previous 6 months shows that subclades within clades 3 and 5 are growing the most rapidly (\FIG{h3n2_tree_dfreq}).
The most rapidly growing subclade is the E62G/R142G/T135K subclade within clade 5.
This analysis is borne out in "local branching index" \cite{neher2014predicting}, which also supports this subclade as a high fitness clade.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree_dfreq.png}
	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by clade growth.}
  The tree, zoomed into clade 3c2.a, is colored by change in clade frequency over the past 6 months.
  }
  \label{h3n2_tree_dfreq}
\end{figure}

\textit{In summary, we think that the clade most likely to increase further in frequency is the E62G/R142G/T135K subclade within clade 5.
However, clade 2 viruses (characterized by N121K/R144K) and clade 3 viruses (characterized by T131K/R142K/R261Q) also appear competitive, although less so than E62G/R142G/T135K viruses.
We expect all three of these clades to contribute to the future H3N2 population.}

>>>>

\clearpage
\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree_with_timetree.png}
  \caption{\textbf{H3N2 / 3c2.a phylogeny colored by genotype.} Colors are chosen to highlight clades 1--5. Clades are in reverse order, that is clade 5 is on top, clade 1 is at the bottom. Each of these clades is non-nested and so all compete against each other. Also shown is a time calibrated phylogeny with equivalent coloring.
  }
  \label{h3n2_tree_with_timetree}
\end{figure}

We observe substantial diversification within the 3c2.a clade over the
past year. There are 5 major subclades within 3c2.a that have reached
appreciable frequency (\FIG{h3n2_tree_with_timetree}). These clades are characterized by the following
mutations and highlighed in the phylogenetic tree in \FIG{h3n2_tree}:

@@

All of these clades have risen in frequency and are now competing with
one another (\FIG{h3n2_mutations}, \FIG{h3n2_mutations_by_region}). Clade 1 (53N, 144R, 171K, 192T, 197H) is highly derived in
terms of sequence, but is at low frequency globally (currently less than
1\%) despite making up the a substantial fraction of 2017 Oceania
viruses. Clade 2 (121K, 144K) had a large upswing around Dec 2016, but has since declined slightly to $\sim$21\% global frequency. Clade 3 (131K, 142K,
261Q) and clade 4 (135K, HA2:150E) have each slowly climbed to
approximately 20\% frequency. However, clade 5 (92R, 311Q) has recently
increased to 39\% in global frequency.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/h3n2_mutations.png}
  \caption{\textbf{Frequency trajectories of 3c2.a subclades partitioned by clade then by region.}
  We estimate frequencies of different clades based on sample counts and collection dates.
  Each clade is estimated according to a characteristic mutation unique to this clade.
  These estimates are based on all available data and global frequencies are weighted according to regional population size and relative sampling intensity .
  We use a Brownian motion process prior to smooth frequencies from month-to-month.
  Transparent bands show an estimate the 95\% confidence interval based on sample counts.
  }
  \label{h3n2_mutations}
\end{figure}

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.95\textwidth]{../figures/sep-2017/h3n2_mutations_by_region.png}
  \caption{\textbf{Frequency trajectories of 3c2.a subclades partitioned by region then by clade.}
  We estimate frequencies of different clades based on sample counts and collection dates.
  Each clade is estimated according to a characteristic mutation unique to this clade.
  These estimates are based on all available data and global frequencies are weighted according to regional population size and relative sampling intensity .
  We use a Brownian motion process prior to smooth frequencies from month-to-month.
  Transparent bands show an estimate the 95\% confidence interval based on sample counts.
  }
  \label{h3n2_mutations_by_region}
\end{figure}

@@

Additionally, an analysis of ``local branching index'' \cite{neher2014predicting} suggests that the E62G/R142G/T135K subclade within clade 5 is the most rapidly growing of circulating viruses and is thus estimated to have the highest fitness.
With the exception of background 3c2.a1 viruses, other circulating 3c2.a are seen to have similar local branching index values.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.8\textwidth]{../figures/sep-2017/h3n2_tree_lbi.png}
	\caption{\textbf{H3N2 / 3c2.a phylogeny colored by local branching index.}
  The tree, zoomed into clade 3c2.a, is colored by local branching index.
  }
  \label{h3n2_tree_dfreq}
\end{figure}

\textit{In summary, we think that the clade most likely to increase further in frequency is the E62G/R142G/T135K subclade within clade 5.
However, clade 2 viruses (characterized by N121K/R144K) and clade 3 viruses (characterized by T131K/R142K/R261Q) also appear competitive (they are present at moderate frequency, have somewhat diverged antigenic phenotypes and are increasing in frequency), although less so than E62G/R142G/T135K viruses (which are growing faster and appear more diverged in antigenic phenotype).
We expect all three of these clades to contribute to the future H3N2 population.}
